Supplementary material

### 2022-01-10 analysis/paper/paper.Rmd
References for supplementary material

# Load processed site and radiocarbon data, and load functions
load(here("analysis/data/derived_data/01data.RData"))
load(here("analysis/data/derived_data/02data.RData"))
load(here("analysis/data/derived_data/06data.RData"))
source(here("analysis/script/04script.R"), local = knitr::knit_global())
```

@@

>>>>

# Load processed site and radiocarbon data, and load functions
load(here("analysis/data/derived_data/01data.RData"))
load(here("analysis/data/derived_data/02data.RData"))
load(here("analysis/data/derived_data/06data.RData"))
source(here("analysis/script/04script.R"), local = knitr::knit_global())

# Exclude Lunaveien and Dybdalshei 2 (See supplementary material)
sites_sa <- filter(sites_sa, !(name %in% c("Dybdalshei 2", "Lunaveien")))

@@

Five of the sites have been associated with farming activities, either in directly or in the form building structures. The first is Nordby 1, which has . The Middle Neolithic phase at Kvastad A2 and Late Neolithic phase at Nauen A are related to farming. Both sites also have ^14^C-dates and lithic inventory associated with Mesolithic forager activities. Finally, @nielsen2021 has recently suggested that Early and Middle Neolithic features from the otherwise younger sites Bratberg and Lang could be related to early agricultural activity in the Oslo fjord region.

### 2022-01-10 analysis/supplementary-materials/supplementary.Rmd
Fixed unpacking of results
References for supplementary material

---
title: "Supplementary material"
author: "Isak Roalkvam"
date: "10/27/2021"
output: pdf_document
bibliography: references.bib
---

```{r,  echo = FALSE, results = 'hide'}
library(here)
library(tidyverse)
library(knitr)
library(kableExtra)

# Read in radiocarbon dates to be plotted alongside analysis for
# each site.
rcarbon <- read.csv(here('analysis/data/raw_data/radiocarbon.csv'))
```

This document shows the simulation results for each individual site analysed in the study with some additional notes pertaining to edits to the DTM, treatment of ^14^C-dates and other relevant information. The sites are presented in alphabetical order.

The first figure shows the result of a test run for the site Hovland 5 where the simulations where run 5000 times to identify when the means for the different distance measures converged. This was used to inform the use of 1000 simulation runs for the subsequent analysis of the other sites, as the means appear to have stabilised sufficiently by this point. Hovland 5 was chosen because the site only has a single ^14^C-date, giving a wide uncertainty range, and because it is located in an area of the landscape where the landscape results in quite large variability on the distance measures.

@@

Subsequent rows for each site then show the simulation results. The first plot on these rows shows the probability density function from which site dates were drawn during simulation. The second plot shows the result of the simulation runs, where the intensity of the colour indicates the number of times the sea was simulated to be present at any given location. The third plot displays a violin plot of the measured distances between site and shore-line across all simulation runs. Measures for central tendency and dispersion associated with these are then given in the table.

```{r adalvestre1,  echo = FALSE, out.width = "500px"}

@@

```{r dalvestre2,  echo = FALSE, out.width = "500px"}
knitr::include_graphics(here("analysis/figures/adalvestre2.png"))

@@

```{r anvik,  echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/anvik.png"))

@@

```{r bakke, echo = FALSE, out.width = "500px"}
knitr::include_graphics(here("analysis/figures/bakke.png"))

@@

```{r dordal, echo = FALSE, out.width = "500px"}

@@

```{r dybdalshei1, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/dybdalshei1.png"))

@@

Dybdalshei 1 was subject to a very limited excavation [@granum2018], and the lack of data might therefore lead one to question whether or not the site actually represent a settlement area. Furthermore, what little lithic material was retrieved is too generic to offer any typological support to the radiocarbon dates of the dated cooking pits. Nonetheless, the site and the ^14^C-dates are still treated the same way as the other sites. This is based on the spatial relation between the lithic artefacts and the features, which offers some additional support to the fact that the cooking pits do in fact represent remnants from visits in the Mesolithic.

```{r dybdalshei2, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/dybdalshei2.png"))

@@

```{r gunnarsrod4, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/gunnarsrod4.png"))

@@

Gunnarsrød 4 [@reitan2014k], and Gunnarsrød 5 (below), were situated close to where the railway runs today, and therefore required editing of the DTM. Based on maps from the report, this appears to have given adequate, although not perfect results [@reitan2014k, fig.14.7.1]. The date from the bottom of the cultural layer was seen as possibly impacted by old-wood effect, a notion that is in line with the simulation results, as this date would have the site located beneath the sea-level [@reitan2014k].

```{r gunnarsrod5, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/gunnarsrod5.png"))

@@

The location of Gunnarsrød 5 is today situated at the exit of a railway tunnel. The construction of the tunnel entrance has lead to a removal of parts of what used to be a hill on which the site was situated [@reitan2014d]. While the editing and interpolation of the DTM appears to give reasonable results for the rest of the flatter marsh area where the railway runs today (see also Gunnarsrød 4, above), the site itself would thus have been situated slightly higher. This has likely led to the negative values on the distance measures. In combination, the simulation results and the site map in the report indicate that the site would not have been situated any significant distance from the shoreline. The values have therefore been set to zero for the final aggregative analysis. A note can also be made on the date that is seen as unrelated to the artefact inventory. This is from charcoal from a pit with ceramics which had a far younger date. The date was therefore interpreted as the result of contamination of some kind.

```{r gunnarsrod6a, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/gunnarsrod6a.png"))

@@

The ^14^C-date to the Stone Age from Gunnarsrød 6a is not related to the lithic inventory on the site, which is clearly Mesolithic [@carrasco2014].

```{r gunnarsrod6b, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/gunnarsrod6b.png"))

@@

As with Gunnarsrød 6a above, the ^14^C-date to the Stone Age from Gunnarsrød 6b is not related to the Mesolithic inventory on the site [@carrasco2014].

```{r gunnarsrod7, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/gunnarsrod7.png"))

@@

Gunnarsrød 7 is situated on a slope, stretching from 55 masl up to 59 [@fossum2014b]. The lithic inventory is of a Middle Mesolithic character and does not match the ^14^C-dates.

```{r gunnarsrod10, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/gunnarsrod10.png"))

@@

Based on the location of Gunnarsrød 10 on a steep ledge, and the site inventory, the ^14^C-date is not believed to be related to the main occupation of the site [@reitan2014l].

```{r hegnavest1, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/hegnavest1.png"))

@@

```{r hegnavest2, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/hegnavest2.png"))

@@

```{r hegnavest3, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/hegnavest3.png"))

@@

```{r hesthagc2, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/hesthagc2.png"))

@@

```{r hesthagc4, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/hesthagc4.png"))

@@

Typology and radiocarbon date match, and there are no modern disturbances in the vicinity of the site [@viken2018b]. The second ^14^C-date is a reference date of charcoal from a context believed not be anthropogenic.

```{r hovland1, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/hovland1.png"))

@@

It was noted in the report that it is uncertain how close to the shoreline Hovland 1 would have been situated, based on the ^14^C-dates [@olsen2013b]. The earliest dates match the lithic inventory. The later date is the only indication of activity from the Early Bronze Age/Early Iron Age. The site is not located far from the highway, which was been edited for the analysis of Hovland 1 and 4, as well as Torstvet (see below).

```{r hovland3, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/hovland3.png"))

@@

```{r hovland4, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/hovland4.png"))

@@

Hovland 4 is well-dated both typologically and with ^14^C-dates to the Mesolithic. It has a similar location to that of Hovland 3 (above), and editing of the highway seems to have been successful.

```{r hovland5, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/hovland5.png"))

@@

```{r hydal4, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/hydal4.png"))

@@

```{r krogenesd1, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/krogenesd1.png"))

@@

All ^14^C-dates from Krøgenes D1 are believed to be related to use of the site, which has been reused multiple times over a long time-span and which is evidenced in the artefact inventory as well [@reitan2018]. To what degree the different dates are related to different or the same settlement phases is difficult to ascertain, but is here, as with for the rest of the sites, simply based on grouping dates overlapping with 99.7% probability. The site is situated around 100 meter from a road to the south, but this does not appear to impact simulation results, and is therefore left unedited in the DTM.

```{r krogenesd2, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/krogenesd2.png"))

@@

```{r kvastada1, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/kvastada1.png"))

@@

```{r kvastada2, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/kvastada2.png"))

@@

```{r kvastada4, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/kvastada4.png"))

@@

While there are a few elements of the lithic inventory from Kvastad A4 that could be related to the Stone Age date, it is not related to main visit to the site, which are Early and Middle Mesolithic in character [@darmark2018c]. The highway runs over the site today, but this is not relevant for the shore-line reconstruction done here.

```{r kvastada9, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/kvastada9.png"))

@@

```{r langangenv1, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/langangenv1.png"))

@@

```{r langangenv3, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/langangenv3.png"))

@@

```{r langangenv5, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/langangenv5.png"))

@@

```{r langangenv6, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/langangenv6.png"))

@@

```{r langangenv7, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/langangenv7.png"))

@@

```{r langemyr, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/langemyr.png"))

@@

```{r lovas1, echo = FALSE, out.width = "500px"}
knitr::include_graphics(here("analysis/figures/lovas1.png"))

@@

```{r lovas2, echo = FALSE, out.width = "500px"}
knitr::include_graphics(here("analysis/figures/lovas2.png"))

@@

```{r lovas3, echo = FALSE, out.width = "500px"}
knitr::include_graphics(here("analysis/figures/lovas3.png"))

@@

```{r lunaveien, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/lunaveien.png"))

@@

```{r nauena, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/nauena.png"))

@@

Nedre Holtan.

```{r nauenc, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/nauenc.png"))

@@

```{r nordby1, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/nordby1.png"))

@@

Building structures from the Late Neolithic, Bronze Age and Early Iron Age [@gjerpe2008b]. Lithics and pottery finds support this. Site located on what is today a highway which was edited also for Nordby 52 (below), but is not relevant for the sea-level reconstruction for Nordby 1.

```{r nordby52, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/nordby52.png"))

@@

```{r pauler1, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/pauler1.png"))

@@

```{r pauler2, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/pauler2.png"))

@@

```{r pjonkerodr1, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/pjonkerodr1.png"))

@@

```{r prestemoen1, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/prestemoen1.png"))

@@

```{r ragnhildrod, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/ragnhildrod.png"))

@@

```{r rodbol54, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/rodbol54.png"))

@@

```{r rognlien, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/rognlien.png"))

@@

```{r sageneb1, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/sageneb1.png"))

@@

```{r sageneb2, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/sageneb2.png"))

@@

```{r solum3, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/solum3.png"))

@@

```{r stokkep1, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/stokkep1.png"))

@@

```{r stokkep5, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/stokkep5.png"))

@@

```{r stokkep8, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/stokkep8.png"))

@@

```{r torstvet, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/torstvet.png"))

@@

```{r trondalen, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/trondalen.png"))

@@

```{r tverdal, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/tverdal.png"))

@@

```{r vallermyrene1a, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/vallermyrene1a.png"))

@@

```{r vallermyrene1b, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/vallermyrene1b.png"))

@@

```{r vallermyrene2, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/vallermyrene2.png"))

@@

```{r vallermyrene4a, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/vallermyrene4a.png"))

@@

```{r vallermyrene4b, echo = FALSE, out.width = "500px"}
include_graphics(here("analysis/figures/vallermyrene4b.png"))

@@

>>>>

---
title: "Supplementary material"
author: "Isak Roalkvam"
date: "10/27/2021"
header-includes:
  - \usepackage{placeins}
output: pdf_document
bibliography: references.bib
---

```{r,  include=FALSE, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(knitr)
library(kableExtra)
knitr::opts_chunk$set(echo = FALSE, cache=FALSE)

# Set up so that I can call floatbarrier using regfloat=TRUE in the chunk 
# headings. See: 
# https://gist.github.com/burchill/8873d2ade156b27e92a238a774ce2758
knitr::knit_hooks$set(plot = function (x, options) {
  float_correct <- function(f, y, opts)  {
    if (is.null(opts$regfloat) || opts$regfloat==FALSE)
      paste0(f(y, opts), "\n\n\\FloatBarrier\n")
    else
      f(y, opts)
  }
  if (!is.null(options$out.width) || !is.null(options$out.height) ||
      !is.null(options$out.extra) || options$fig.align != "default" ||
      !is.null(options$fig.subcap)) {
    if (is.null(options$fig.scap))
      options$fig.scap = NA
    return(float_correct(knitr:::hook_plot_tex, x, options))
  }
  return(float_correct(knitr:::hook_plot_md_base, x, options))
})
```

```{r}
# Read in radiocarbon dates to be plotted alongside analysis for
# each site.
rcarbon <- read.csv(here('analysis/data/raw_data/radiocarbon.csv'))
```

This document shows the simulation results for each individual site analysed in the study with some additional notes pertaining to edits to the DTM, treatment of ^14^C-dates and other relevant information. The sites are presented in alphabetical order. Tables holding data on excavated sites and radiocarbon dates, and spatial data for site limits and site features are available in the online repoistory for the paper at https://osf.io/7F9SU

Sites with ^14^C-dates to the Stone Age, but for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975], Nedre Holtan, Rugtvedt [@odgaard1994], Auve [@østmo2008] and Torsrød [@østmo1975]. These sites are therefore not included in the analysis.

The first figure shows the result of a test run for the site Hovland 5, where the simulations were run 5000 times to identify when the means for the different distance measures converged. This was used to inform the use of 1000 simulation runs for the subsequent analysis of the other sites, as the means appear to have stabilised sufficiently by this point. Hovland 5 was chosen because the site only has a single ^14^C-date, giving a wide uncertainty range, and because it is located in an area of the landscape where the landscape results in quite large variability on the distance measures.

@@

Subsequent rows for each site then show the simulation results. The first plot on these rows shows the probability density function from which site dates were drawn during simulation. The second plot shows the result of the simulation runs, where the intensity of the colour indicates the number of times the sea was simulated to be present at any given location. The third plot displays a violin plot of the measured distances between site and shore-line across all simulation runs.

```{r adalvestre1,  echo = FALSE, out.width = "500px", regfloat=TRUE}

@@

```{r dalvestre2,  echo = FALSE, out.width = "500px", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/adalvestre2.png"))

@@

```{r anvik,  echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/anvik.png"))

@@

```{r bakke, echo = FALSE, out.width = "500px", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/bakke.png"))

@@

```{r dordal, echo = FALSE, out.width = "500px", regfloat=TRUE}

@@

```{r dybdalshei1, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/dybdalshei1.png"))

@@

Dybdalshei 1 was subject to a very limited excavation [@granum2018], and the lack of data might therefore lead one to question whether or not the site actually represent a settlement area. Furthermore, what little lithic material was retrieved is too generic to offer any typological support to the radiocarbon dates of the dated cooking pits. Nonetheless, the site and the ^14^C-dates are still treated the same way as the other sites. This is based on the spatial relation between the lithic artefacts and the features, which offers some additional support to the notion that the cooking pits represent remnants from visits in the Mesolithic.

```{r dybdalshei2, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/dybdalshei2.png"))

@@

```{r gunnarsrod4, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/gunnarsrod4.png"))

@@

The rock-shelter site Gunnarsrød 4 [@reitan2014k], and Gunnarsrød 5 (below), were situated close to where the railway runs today, and therefore required editing of the DTM. Based on maps from the report, this appears to have given adequate, although not perfect results [@reitan2014k, fig.14.7.1]. The date from the bottom of the cultural layer was seen as possibly impacted by old-wood effect, a notion that is in line with the simulation results, as this date would have the site located beneath the sea-level [@reitan2014k].

```{r gunnarsrod5, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/gunnarsrod5.png"))

@@

The location of Gunnarsrød 5 is today situated at the exit of a railway tunnel. The construction of the tunnel entrance has lead to a removal of parts of what used to be a hill on which the site was situated [@reitan2014d]. While the editing and interpolation of the DTM appears to give reasonable results for the rest of the flatter marsh area where the railway runs today (see also Gunnarsrød 4, above), the site itself would thus have been situated slightly higher. This has likely led to the negative values on the distance measures. In combination, the simulation results and the site map in the report [@reitan2014d, fig.9.1] indicate that the site would not have been situated any significant distance from the shoreline. The values have therefore been set to zero for the final aggregative analysis. A note can also be made on the date that is seen as unrelated to the artefact inventory. This is from charcoal from a pit with ceramics which had a far younger date. The date was therefore interpreted as the result of contamination of some kind.

```{r gunnarsrod6a, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/gunnarsrod6a.png"))

@@

The ^14^C-date to the Neolithic from Gunnarsrød 6a is not related to the lithic inventory on the site, which is clearly Mesolithic [@carrasco2014].

```{r gunnarsrod6b, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/gunnarsrod6b.png"))

@@

As with Gunnarsrød 6a above, the ^14^C-date to the Neolithic from Gunnarsrød 6b is not related to the Mesolithic inventory on the site [@carrasco2014].

```{r gunnarsrod7, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/gunnarsrod7.png"))

@@

 The lithic inventory from Gunnarsrød 7 is of a Middle Mesolithic character and does not match the ^14^C-dates [@fossum2014b].

```{r gunnarsrod10, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/gunnarsrod10.png"))

@@

Based on the location of Gunnarsrød 10 on a steep ledge indicating a shore-bound location, and the site inventory, the ^14^C-date is not believed to be related to the main occupation of the site [@reitan2014l].

```{r hegnavest1, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/hegnavest1.png"))

@@

```{r hegnavest2, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/hegnavest2.png"))

@@

```{r hegnavest3, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/hegnavest3.png"))

@@

```{r hesthagc2, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/hesthagc2.png"))

@@

```{r hesthagc4, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/hesthagc4.png"))

@@

Typology and radiocarbon date match, and there are no modern disturbances in the vicinity of Hesthag C4 [@viken2018b]. The second ^14^C-date is a reference date of charcoal from a context believed not be anthropogenic.

```{r hovland1, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/hovland1.png"))

@@

It was noted in the report that it is uncertain how close to the shoreline Hovland 1 would have been situated, based on the ^14^C-dates [@olsen2013b]. The earliest dates match the lithic inventory. The later date is the only indication of activity from the Early Bronze Age/Early Iron Age. The site is not located far from the highway, which was also edited for the analysis of Hovland 1 and 4, as well as Torstvet (see below).

```{r hovland3, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/hovland3.png"))

@@

```{r hovland4, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/hovland4.png"))

@@

Hovland 4 is well-dated both typologically and with ^14^C-dates to the Mesolithic [@mansrud2013a]. It has a similar location to that of Hovland 3 (above), and editing of the highway seems to have been successful.

```{r hovland5, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/hovland5.png"))

@@

```{r hydal4, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/hydal4.png"))

@@

```{r krogenesd1, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/krogenesd1.png"))

@@

All ^14^C-dates from Krøgenes D1 are believed to be related to use of the site, which has been reused multiple times over a long time-span [@reitan2018]. This is evidenced by the artefact inventory as well. To what degree the different dates are related to different or the same settlement phases is difficult to ascertain, but is here, as with for the rest of the sites, simply based on grouping dates overlapping with 99.7% probability. The site is situated around 100 meter from a road to the south, but this does not appear to impact simulation results, and is therefore left unedited in the DTM.

```{r krogenesd2, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/krogenesd2.png"))

@@

```{r kvastada1, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/kvastada1.png"))

@@

```{r kvastada2, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/kvastada2.png"))

@@

```{r kvastada4, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/kvastada4.png"))

@@

While there are a few elements of the lithic inventory from Kvastad A4 that could be related to the late Middle Mesolithic ^14^C-date, it is not related to the main visits to the site, which are Early Mesolithic in character [@darmark2018c]. The highway runs over the site today, but this is not relevant for the shore-line reconstruction done here.

```{r kvastada9, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/kvastada9.png"))

@@

```{r langangenv1, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/langangenv1.png"))

@@

```{r langangenv3, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/langangenv3.png"))

@@

```{r langangenv5, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/langangenv5.png"))

@@

```{r langangenv6, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/langangenv6.png"))

@@

```{r langangenv7, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/langangenv7.png"))

@@

```{r langemyr, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/langemyr.png"))

@@

```{r lovas1, echo = FALSE, out.width = "500px", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/lovas1.png"))

@@

```{r lovas2, echo = FALSE, out.width = "500px", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/lovas2.png"))

@@

```{r lovas3, echo = FALSE, out.width = "500px", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/lovas3.png"))

@@

```{r lunaveien, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/lunaveien.png"))

@@

```{r nauena, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/nauena.png"))

@@

```{r nauenc, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/nauenc.png"))

@@

```{r nordby1, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/nordby1.png"))

@@

Site dated to the Late Neolithic, Bronze Age and Early Iron Age [@gjerpe2008b] with long-house 4 dated to the LN. Lithics and pottery finds support the dates. Site located on what is today a highway which was edited also for Nordby 52 (below), but is not relevant for the sea-level reconstruction for Nordby 1.

```{r nordby52, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/nordby52.png"))

@@

```{r pauler1, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/pauler1.png"))

@@

```{r pauler2, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/pauler2.png"))

@@

```{r pjonkerodr1, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/pjonkerodr1.png"))

@@

```{r prestemoen1, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/prestemoen1.png"))

@@

```{r ragnhildrod, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/ragnhildrod.png"))

@@

```{r rodbol54, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/rodbol54.png"))

@@

```{r rognlien, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/rognlien.png"))

@@

```{r sageneb1, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/sageneb1.png"))

@@

```{r sageneb2, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/sageneb2.png"))

@@

```{r solum3, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/solum3.png"))

@@

```{r stokkep1, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/stokkep1.png"))

@@

```{r stokkep5, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/stokkep5.png"))

@@

```{r stokkep8, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/stokkep8.png"))

@@

```{r torstvet, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/torstvet.png"))

@@

```{r trondalen, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/trondalen.png"))

@@

```{r tverdal, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/tverdal.png"))

@@

```{r vallermyrene1a, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/vallermyrene1a.png"))

@@

```{r vallermyrene1b, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/vallermyrene1b.png"))

@@

```{r vallermyrene2, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/vallermyrene2.png"))

@@

```{r vallermyrene4a, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/vallermyrene4a.png"))

@@

```{r vallermyrene4b, echo = FALSE, out.width = "500px", regfloat=TRUE}
include_graphics(here("analysis/figures/vallermyrene4b.png"))

@@

Bratsberg was originally two sites [@wenn2012]. [@nielsen2021] has interpreted this as an indicator of early farming.
